### 2025-02-07 paper/body_ecml.tex
slowly slowly

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CE to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant. For our intents and
purposes, counterfactual training aims at improving model explainability
by aligning models with counterfactuals that meet both desiderata,
plausibility and actionability. Formally, we define explainability as
follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
counterfactual explanation for that target, similar to how conditional
sampling can be used for JEMs \cite{grathwohl2020your}. Intuitively, the
gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while at same time increasing the
energy of counterfactuals (negative samples) \cite{du2019implicit}. As
the generated counterfactuals get more plausible
(Definition~\ref{def-explainability}) over the cause of training, these
two opposing effects gradually balance each out \cite{lippe2024uvadlc}.

@@

\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}

>>>>

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CE to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(y^+ \sim p(y)\) and then generating a counterfactual
explanation for that target, similar to how conditional sampling is used
to draw from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class.

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
\cite{du2019implicit}. As the generated counterfactuals get more
plausible (Definition~\ref{def-explainability}) over the cause of
training, these two opposing effects gradually balance each out
\cite{lippe2024uvadlc}.

@@

\subsection{Encoding Actionability
Constraints}\label{encoding-actionability-constraints}

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
\cite{altmeyer2023explaining}. In this context, domain constraints can
be imposed by simply projecting counterfactuals back to the specified
domain, if the previous gradient step resulted in updated feature values
that were out-of-domain. Mutability constraints can similarly be
enforced by setting partial derivatives to zero to ensure that features
are only mutated in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. This is best
explained through a simple example involving a linear classifier:
\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
\(y^+ \sim p(y)\) we have:

\begin{equation}\phantomsection\label{eq-grad}{
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
}\end{equation}

In other words, contrastive divergence will push down (up) on the
coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
(higher) for counterfactuals than samples. This has the intended effect
of aligning counterfactual explanations with the data, as discussed
above.

Crucially though, we want to avoid penalizing implausibility that arises
due to mutability constraints. Suppose, for example, that feature \(d\)
represents \emph{age} and our model predicts the likelihood of credit
default.

### 2025-02-07 paper/preprint/paper.tex
slowly slowly

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant. For our intents and
purposes, counterfactual training aims at improving model explainability
by aligning models with counterfactuals that meet both desiderata,
plausibility and actionability. Formally, we define explainability as
follows:

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
counterfactual explanation for that target, similar to how conditional
sampling can be used for JEMs
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). Intuitively,
the gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while at same time increasing the
energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the cause of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
In this context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step moved resulted in updated feature values out of the domain.
Mutability constraints can similarly be enforced by simply clipping
gradients to ensure that features are only mutated in the allowed
direction, if at all.

Since both types of constraints directly affect \(\mathbf{x}^\prime\),
they also influence model outcomes through Equation~\ref{eq-div}.

Mutability constraints

@@

>>>>

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(y^+ \sim p(y)\) and then generating a counterfactual
explanation for that target, similar to how conditional sampling is used
to draw from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}). In particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class.

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the cause of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
In this context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Mutability constraints can similarly be enforced by setting partial
derivatives to zero to ensure that features are only mutated in the
allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. This is best
explained through a simple example involving a linear classifier:
\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
\(y^+ \sim p(y)\) we have:

\begin{equation}\phantomsection\label{eq-grad}{
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
}\end{equation}

In other words, contrastive divergence will push down (up) on the
coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
(higher) for counterfactuals than samples. This has the intended effect
of aligning counterfactual explanations with the data, as discussed
above.

Crucially though, we want to avoid penalizing implausibility that arises
due to mutability constraints. Suppose, for example, that feature \(d\)
represents \emph{age} and our model predicts the likelihood of credit
default.

@@

\section{Notation}\label{notation}

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\end{itemize}

### 2025-02-07 paper/sections/method.qmd
slowly slowly

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements. In the context of CE, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CE to provide algorithmic recourse (AR) to individuals affected by opaque models in practice. This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant. For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability. Formally, we define explainability as follows:

@@

$$
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
$$ {#eq-obj}

@@

$$
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
$$ {#eq-div}

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $\mathbf{y}^+ \sim p(\mathbf{y})$ and then generating a counterfactual explanation for that target, similar to how conditional sampling can be used for JEMs [@grathwohl2020your]. Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

@@

## Encoding Domain Knowledge

>>>>

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements. In the context of CE, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CE to provide algorithmic recourse (AR) to individuals affected by opaque models in practice. This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a mutable feature. For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability. Formally, we define explainability as follows:

@@

$$
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
$$ {#eq-obj}

@@

$$
\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
$$ {#eq-div}

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target, similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your. In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the target class.

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

@@

## Encoding Actionability Constraints

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only mutated in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. This is best explained through a simple example involving a linear classifier: $\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}$. Then for $y^+ \sim p(y)$ we have:

$$
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
$$ {#eq-grad}

In other words, contrastive divergence will push down (up) on the coefficient $\theta[d]$ corresponding to a feature $d$ that is lower (higher) for counterfactuals than samples. This has the intended effect of aligning counterfactual explanations with the data, as discussed above.

Crucially though, we want to avoid penalizing implausibility that arises due to mutability constraints. Suppose, for example, that feature $d$ represents *age* and our model predicts the likelihood of credit default.
